---
title: "Geography 13"
author: "[Jason Allen](https://jcallen1995.github.io/)"
subtitle: "Lab 02: COVID-19 Pandemic"
output:
  html_document:
    code_folding: 'hide'
    theme: journal

---

```{r, echo=TRUE, include=FALSE} 
#this code chunk is for setting up the necessary data and variables
library(tidyverse)
library(knitr)
library(readxl)
library(zoo)
library(kableExtra)
library(readr)
library(maps)
url <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
COVID <- read_csv(url)
target_state <- "California"
#for reading in population data from a file in data
#so, because the current working directory of this chunk is the /docs folder
#I have to establish an absolute path since the pop estimate is in the data
#folder which is in a different folder in geog-13-labs
Population_Estimates = read_excel("../data/PopulationEstimates.xls")
#selection of state for question 2
facet_states = c("New York", "California", "Louisiana", "Florida")



```


```{r, echo=TRUE, include=FALSE}
#this code chunk is for filtering our raw data into a usable manner
Analyzed_Data <- COVID %>% 
  filter(state == target_state) %>% 
  group_by(county) %>% 
  mutate(new_cases = cases - lag(cases)) %>% 
  ungroup()
Daily_Data <- Analyzed_Data %>% 
  filter(date == max(date))
```

```{r, echo=TRUE}
#This code chunk is for making and displaying knitr tables
Most_County_Cum_Cases <- Daily_Data %>% 
  slice_max(cases, n = 5) %>% 
  select(county, cases)
Most_County_Daily_Cases <- Daily_Data %>% 
  slice_max(new_cases, n = 5) %>% 
  filter(new_cases > 0) %>% 
  arrange(-new_cases) %>%
  select(county, new_cases)
print_cum <- Most_County_Cum_Cases %>% #print a nice table for daily, cumulative cases
  kable(col.names = c("County", "Cases"),
        caption = "Daily Cumulative Cases",
        format.args = list(big.mark = ',')) %>% 
  kable_styling("striped", full_width = TRUE, font_size = 16)
(print_cum)
print_daily <- Most_County_Daily_Cases %>% #print a nice table for daily cases
  kable(col.names = c("County", "Cases"),
        caption = "Daily Cases",
        format.args = list(big.mark = ',')) %>% 
  kable_styling("striped", full_width = TRUE, font_size = 16)
(print_daily)
```

```{r, echo=TRUE, include=FALSE}
#new code chunk to join the population estimate table to the covid data
#rename the insane paragraph long fips column name
names(Population_Estimates)[1] <- "fips"
#join pop estimates and daily data based on fips number
Cum_Cases_Per_Cap <- Analyzed_Data %>% 
  filter(date == max(date)) %>% 
  left_join(Population_Estimates, by = c("fips" = "fips")) %>% 
  #add a cases per capita column by dividing cases by population in 2019 census
  mutate(cum_cases_per_cap = cases / as.numeric(...20)) %>% 
  slice_max(cum_cases_per_cap, n = 5) %>% 
  select(county, cum_cases_per_cap)
#join pop estimates and daily data based on fips number
New_Cases_Per_Cap <- Daily_Data %>% 
  left_join(Population_Estimates, by = c("fips" = "fips")) %>% 
  #add a cases per capita column by dividing cases by population in 2019 census
  mutate(new_cases_per_cap = new_cases / as.numeric(...20)) %>% 
  slice_max(new_cases_per_cap, n = 5) %>% 
  select(county, new_cases_per_cap)
```



```{r, echo=TRUE, message=FALSE}
#this code chunk if for getting the last 14 days average
Average_14_Day <- Analyzed_Data %>% 
  left_join(Population_Estimates, by = c("fips" = "fips")) %>% 
  filter(date > max(date) - 14) %>% 
  group_by(state, date) %>% 
  summarize(total_cases_per_100k = sum(cases, na.rm = TRUE) / sum(as.numeric(...20), na.rm = TRUE) * 100000) %>% 
  ungroup() %>% 
  group_by(state) %>% 
  summarize(total_14_day_average = sum(total_cases_per_100k / 14))
print_avg <- Average_14_Day %>% 
  kable(col.names = c("State", "Cases Per 100k"),
        caption = "Last 14 day average of cases per 100,000 people") %>% 
  kable_material_dark() %>% 
  kable_styling(full_width = FALSE, font_size = 20, position = "left")
(print_avg)
```


```{r, echo=TRUE, include=FALSE}
#calculating the variables needed for the report
#total cumulative cases
total_state_cases <- Daily_Data %>% 
  group_by(state) %>% 
  summarise(total = sum(cases, na.rm = TRUE)) %>% 
  ungroup()
total_cases = as.numeric(total_state_cases[2])
#total new cases
total_new_state_cases <- Daily_Data %>% 
  group_by(state) %>% 
  summarise(total = sum(new_cases, na.rm = TRUE)) %>% 
  ungroup()
total_new_cases = as.numeric(total_new_state_cases[2])
#for printing the current date
recent_date <- as.Date(Daily_Data$date[1])
#cast as character cause rmarkdown prints numerics in scientific notation
tot_cases = as.character(total_cases)
#total safe counties
#redo the whole average 14 day calculation without summarizing to state
Average_14_Day_County <- Analyzed_Data %>% 
  left_join(Population_Estimates, by = c("fips" = "fips")) %>% 
  filter(date > max(date) - 14) %>% 
  group_by(county, date) %>% 
  summarize(total_cases_per_100k_county = sum(new_cases, na.rm = TRUE) / sum(as.numeric(...20), na.rm = TRUE) * 100000) %>% 
  ungroup() %>% 
  group_by(county) %>% 
  summarize(total_14_day_average_county = sum(total_cases_per_100k_county / 14)) %>% 
  ungroup() %>% 
  filter(total_14_day_average_county < 100)
#number of safe counties
safe_counties = nrow(Average_14_Day_County)
```

****

<br><br>


# The total number of cases in `r (target_state)`is `r (tot_cases)` on `r (recent_date)`.    
# The total number of new cases on `r (recent_date)` is `r (total_new_cases)`.  
# The total number of safe counties in `r (target_state)` is `r (safe_counties)`.  
  
****
<br><br>
  
#### Now let's look at 4 states comparing new cases.

```{r, echo=TRUE, include=FALSE, message=FALSE}
#group by state and select specific states
New_State_Cases <- COVID %>% 
  filter(state == c(facet_states)) %>% 
  filter(county != "Unknown") %>% 
  group_by(fips) %>% 
  mutate(new_cases = cases - lag(cases)) %>% 
  ungroup() %>% 
  filter(new_cases != "NA") %>% 
  group_by(state, date) %>% 
  summarise(total_daily_state_cases = sum(new_cases)) %>% 
  ungroup() %>% 
  mutate(roll_7 = rollmean(total_daily_state_cases, 7,fill = NA, align = "right")) %>% 
  filter(roll_7 != "NA") 
```

```{r, echo=TRUE, message=FALSE}
#this chunk if for ggplotting the first set of facets
New_State_Cases %>% 
  ggplot(aes(x = date, y = roll_7))+
  geom_col(col = "red")+
  facet_wrap(~state)+
  labs(title = "State Daily COVID Cases",
       x = "Date",
       y = "Cases",
       subtitle = "Data from NY Times COVID cumulative count")
```
  
****  
<br> 

### This can be misleading, so let's look at the data when ajusted by total population.

```{r echo=TRUE, message=FALSE, results='hide', warning=FALSE}
#redo new state case and plot without doing the roll 7
New_State_Cases_Per_Cap <- COVID %>% 
  filter(state == c(facet_states)) %>% 
  filter(county != "Unknown") %>% 
  group_by(fips) %>% 
  mutate(new_cases = cases - lag(cases)) %>% 
  ungroup() %>% 
  filter(new_cases != "NA") %>% 
  group_by(state, date) %>% 
  summarise(total_daily_state_cases = sum(new_cases)) %>% 
  ungroup()
Manip_Pop_Raw <- Population_Estimates %>% 
  right_join(New_State_Cases_Per_Cap, by = c("...3" = "state")) %>% 
  filter(total_daily_state_cases != "NA") %>% 
  select(date, ...3, total_daily_state_cases, ...20) %>% 
  mutate(avg_per_cap = total_daily_state_cases / as.numeric(...20)) %>% 
  ggplot(aes(x = date, y = avg_per_cap))+
  geom_col(col = "gold")+
  facet_wrap(~...3)+
  labs(title = "State Daily COVID Cases Per Capita",
       x = "Date",
       y = "Cases Per Cap",
       subtitle = "Data from NY Times COVID cumulative count")
(Manip_Pop_Raw)
  
```
<br><br>  
  
#### This is helpful, but finding the 7 day average will be more comparable to our previous plot.

```{r, eval=TRUE, echo=TRUE, message=FALSE}
#this code chunk is about doin stuff with the population and what now. per kappita, im tired
#let's manipulate population estimate first
Manip_Pop <- Population_Estimates %>% 
  right_join(New_State_Cases, by = c("...3" = "state")) %>% 
  filter(roll_7 != "NA") %>% 
  select(date, ...3, roll_7, ...20) %>% 
  mutate(avg_per_cap = roll_7 / as.numeric(...20)) %>% 
  ggplot(aes(x = date, y = avg_per_cap))+
  geom_col(col = "gold")+
  facet_wrap(~...3)+
  labs(title = "State Daily COVID Cases Per 7 Day Average Capita",
       x = "Date",
       y = "Cases Per Cap Avg",
       subtitle = "Data from NY Times COVID cumulative count")
(Manip_Pop)
  
```
<br><br>  
  
### As can be seen through this second plot what takes into account the number of cases
### in relation to the total population, the rates are much more comparable.  
 
****
  
  
<br><br><br>  
  
## This plot is for the weighted average of COVID cases across the US.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Extra credit attempt: weighted mean center
#import spatial data
County_Centroids <- read_csv("../data/county-centroids.csv")
#Start by adding geospatial data to covid data via fips
Geo_COVID_Data <- COVID %>% 
  left_join(County_Centroids, by = c("fips" = "fips")) %>% 
  filter(LON != "NA") %>% 
  filter(county != "Unknown") %>% 
  group_by(date) %>% 
  summarise(weighted_mean_X = sum(LON * cases, na.rm = TRUE) / sum(cases, na.rm = TRUE), 
            weighted_mean_Y = sum(LAT * cases, na.rm = TRUE) / sum(cases, na.rm = TRUE),
            daily_cases_total = sum(cases, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(m = as.factor(format(date, "%m")))

ggplot() +
  borders("state", fill = "gray90", colour = "white") +
  geom_point(data = Geo_COVID_Data, aes(x = weighted_mean_X, y = weighted_mean_Y, col = m, size = daily_cases_total))+
  labs(title = "Weighted Mean of US COVID Cases",
     x = "Longitude",
     y = "Latitude",
     subtitle = "US Boundaries Data",
     color = "Month",
     size = "Cases")
  

```
  
  
